---
title: "Summarise taxonmic distribution in the EMP dataset"
author: 
  name: "Roey Angel"
  email: "roey.angel@bc.cas.cz"
  affiliation: "Biology Centre CAS, Czechia"
  github-repo: "https://github.com/roey-angel"
  url: 'https://soilwater.eu/people/angel-roey/'
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: references.bib
link-citations: yes
always_allow_html: true
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 5
    df_print: "kable"
    keep_html: TRUE
---

```{r libraries, include=F}
library(extrafont) # Tools for using fonts, CRAN v0.17
library(tidyverse) # Easily Install and Load the 'Tidyverse', CRAN v1.3.0 
library(scales) # Scale Functions for Visualization, CRAN v1.1.1
library(ggtext) # Improved Text Rendering Support for 'ggplot2', CRAN v0.1.0
library(see) # Visualisation Toolbox for 'easystats' and Extra Geoms, Themes and Color Palettes for 'ggplot2', CRAN v0.6.0
library(svglite) # An 'SVG' Graphics Device, CRAN v1.2.3.2 
library(viridis) # Default Color Maps from 'matplotlib'
library(phyloseq) # Handling and analysis of high-throughput microbiome census data, Bioconductor v1.32.0 
library(speedyseq) # Faster implementations of phyloseq functions, [github::mikemc/speedyseq] v0.5.3.9001 
library(patchwork) # The Composer of Plots 
library(MicrobiomeR) # Analyze Microbiome Data, Bioconductor v0.7.0 
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax # CRAN v1.3.1 
library(metacoder) # Tools for Parsing, Manipulating, and Graphing Taxonomic Abundance Data  # [::NA/NA] v0.3.4 # CRAN v0.3.4 
```

```{r style settings, echo=F, message=F, warning=F, results="asis"}
options(width = 90, 
        knitr.table.format = "html") 
knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE, 
  message = FALSE,
  cache.lazy = FALSE,
  dev = c("png"),
  dev.args = list(png = list(bg = 'white', type = "cairo")),
  dpi = 150,
  cache.path = "EMP_cache/",
  fig.path = "EMP_figures/"
)
```

```{r functions, include=F}
PS_merge_samples <- function(ps = Ps_obj, grouping_name = "empo_3", fun = "mean") {
  # Like phyloseq::merge_samples() but retains chr and fct information in sample_data()

  # Merge/agglomerate the sample indices of a phyloseq object according to a categorical variable contained in a sample_data or a provided factor.
  # Unlike merge_samples() this function will only apply a `mean` to numeric variables in `samples_data()` and will retain all unique values of any non-numeric variable.
  # chr or fct variables with non-identical levels in the marged samples are combined to a comma-separated string.
  require(dplyr)
  require(purrr)
  require(phyloseq)

  if (taxa_are_rows(ps)) {ps <- t(ps)} # needs to be in sample-by-species orientation
  SD2merge <- as_tibble(sample_data(ps)) # grab sample_data
  org_col_names <- colnames(SD2merge) # retain original sample_data variable order
  grouping_col <- select(SD2merge, group = all_of(grouping_name)) # grab grouping var
  # grap factor variables
  SD2merge %>%
    select_if(is.factor) %>%
    colnames() ->
    fct_vars

  # merge the OTU table
  ps %>%
    otu_table() %>%
    as(., "matrix") %>%
    rowsum(., as_vector(grouping_col)) %>%
    # convert back to otu_table, and return
    otu_table(., taxa_are_rows = FALSE) ->
    merged_OTU_table

  # merge numeric
  SD2merge %>%
    select_if(is.numeric) %>%
    bind_cols(grouping_col, .) %>%
    group_by(group) %>%
    summarise_all(., fun, na.rm = TRUE) ->
    numeric_cols

  # merge other
  SD2merge %>%
    select_if(negate(is.numeric)) %>%
    bind_cols(grouping_col, .) %>%
    group_by(group) %>%
    summarise_all(list(~paste(unique(.), collapse = ",")))  ->
    other_cols

  # join merged numeric and other columns
  full_join(numeric_cols, other_cols, by = "group") %>%
    select(-group) %>%
    select(all_of(org_col_names)) %>% # order columns like they were
    mutate_at(fct_vars, ~(factor(.))) %>% # return factor type to fct vars
    mutate(torownames = !!sym(grouping_name)) %>% 
    column_to_rownames(var = "torownames") ->
    merged_SD

  # build and return a phyloseq object
  return(phyloseq(otu_table(merged_OTU_table),
                  sample_data(merged_SD),
                  tax_table(ps)))
}
```

## Summarise the taxonmic distribution across different sample types in the Earth Microbiome Project
The Earth Microbiome Project ([EMP](1)) is collaborative effort to curate and characterise microbial taxonomy and function on Earth [@thompson_communal_2017].  
The aim of this script is to summarise the taxonmic distribution across different sample types, as defined by the EMP. The script summarises phyla by default but can also summarise the sequences at different taxonimc levels. The script uses the 3^rd^ calssification level of the [EMP Ontology](3) (`empo_3`), but this can also be altered.

### Set main parameters and download data tables from the EMP FTP site

```{r parameters}
ftp <- "ftp://ftp.microbio.me/emp/release1/"
biom_file <- "emp_cr_silva_16S_123.subset_10k.biom"
map_file <- "emp_qiime_mapping_subset_10k.tsv"
new_map_file <- "emp_qiime_mapping_subset_10k_noBLNK.csv"
Rank <- "Phylum"
Grouping_var <- "empo_3"
EMPO_levels2keep <- c("Soil (non-saline)", "Plant rhizosphere", "Animal corpus", "Water (saline)", "Sediment (saline)", "Water (non-saline)", "Hypersaline (saline)", "Sediment (non-saline)", "Animal distal gut", "Animal proximal gut", "Aerosol (non-saline)", "Animal secretion", "Animal surface", "Plant corpus", "Surface (non-saline)", "Plant surface", "Surface (saline)") # remove unwanted environments
```

```{r download files}
biom_ftp <- paste0(ftp, "otu_tables/closed_ref_silva/", biom_file)
map_ftp <- paste0(ftp, "mapping_files/", map_file)
download.file(biom_ftp, biom_file)
download.file(map_ftp, map_file)

# fill-in missing values with NA, convert to CSV
system("awk 'BEGIN { FS = OFS = \"\t\" } { for(i=1; i<=NF; i++) if($i ~ /^ *$/) $i = \"NA\" }; 1' emp_qiime_mapping_subset_2k.tsv  |  awk 'BEGIN { FS=\",\"; OFS=\";\" } {$1=$1; print}' |  awk 'BEGIN { FS=\"\t\"; OFS=\",\" } {$1=$1; print}' > emp_qiime_mapping_subset_10k_noBLNK.csv")
```

### Generate a phyloseq object from the biom and (modified) map files
```{r generate phyloseq obj}
(EMP_otu_tax <- import_biom(biom_file, parseFunction = parse_taxonomy_silva_128, 
                            parallel = TRUE))
read_csv(new_map_file,
                    trim_ws = TRUE) %>% 
  column_to_rownames("#SampleID") %>% 
  sample_data() ->
  EMP_map

EMP <- merge_phyloseq(EMP_otu_tax, EMP_map)

# for debugging and testing
# EMP %>%
#   subset_samples(empo_3 %in% EMPO_levels2keep) %>%
#   prune_samples(sample_names(.)[sample(1:nsamples(.), 100)], .) %>%
#   phyloseq::filter_taxa(., function(x) sum(x) > 0, TRUE) %>%
#   prune_taxa(taxa_names(.)[sample(1:ntaxa(.), 1000)], .) ->
#   EMP_subset

EMP %>% 
  PS_merge_samples(grouping_name = Grouping_var, fun = "mean") ->
  EMP_merge
```

### Generate a summary table of taxonomic distributions across sample types

```{r summarise phyla}
# summarise abundance by environment and phylum
EMP_merge %>%
  tax_glom(taxrank = Rank) %>%                     # agglomerate at 'Rank' level
  psmelt() %>%                                         # Melt to long format
  arrange(Rank)  %>%                                  # arrange by 'Rank'
  group_by_at(c(Grouping_var, Rank)) %>% 
  # summarise and create a column with the relative abundance
  summarise(Abundance = sum(Abundance)) %>%
  mutate(Rreq = paste0(round(100 * Abundance / sum(Abundance), 2), "%")) ->
  taxa_sum_df

# determine the phylum order by decreasing abundance (for plotting)
EMP_merge %>%
  tax_glom(taxrank = Rank) %>%                     # agglomerate at 'Rank' level
    transform_sample_counts(., function(x) x / sum(x) ) %>% 
  psmelt() %>%                                        # Melt to long format
  arrange(Rank)  %>%                                  # arrange by 'Rank'
  group_by_at(c("OTU", Rank)) %>% 
  summarise(Abundance = sum(Abundance)) %>%
  arrange(Abundance) ->
  Rank_order 

# print summary table
taxa_sum_df %>% 
  filter(Abundance != 0) %>%
  kable(., digits = c( 0, 1)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)

# save summary table
taxa_sum_df %>% 
  filter(Abundance != 0) %>% 
  write.csv(file = paste0("EMP_", Grouping_var,"_", Rank,".csv"))
```

### Plot a heatmap
```{r plot heatmanp, fig.width=18, fig.height=25}
# Agglomerate taxa
EMP_merge %>% 
  tax_glom(Rank, NArm = TRUE) %>% 
  transform_sample_counts(function(x) x / sum(x) * 100) ->
  Ps_obj_filt_glom

plot_heatmap(
  Ps_obj_filt_glom,
  # method = "NMDS",
  # distance = "bray",
  trans = log_trans(4),
  sample.order = Grouping_var,
  sample.label = Grouping_var,
  taxa.label = Rank,
  taxa.order = Rank_order$OTU
) + 
  theme_bw(base_size = 18) + 
  theme(axis.text.x = element_text(angle = 45.0, vjust = 1, hjust = 1)) +
  scale_fill_viridis(trans = "log", labels = scales::number_format(accuracy = 0.1), option = "magma", na.value = "black") +
  labs(fill = "Abundance (%)")
```

### Plot a bar chart
```{r plot bar, fig.width=18, fig.height=30}
EMP_merge %>% 
  tax_glom(., Rank) %>% 
  transform_sample_counts(., function(x) x / sum(x) * 100) %>% 
  psmelt() %>%  
  mutate_if(is.character, as.factor) %>% 
  mutate(Phylum = fct_relevel(pull(., Rank), rev(unique(pull(Rank_order, Rank))))) %>% 
  ggplot(aes(fill = !!sym(Rank), x = !!sym(Grouping_var), y = log10(Abundance))) +
  geom_col() +
  theme_set(theme_bw()) +
  theme(axis.text.x = element_text(angle = 45.0, vjust = 1, hjust = 1),
        legend.position = "bottom") +
  labs(y = "Log10 rel. abundance") +
  facet_wrap(~get(Rank), ncol = 5) 
```

### Plot a violin and dot chart
```{r plot violin, fig.width=12, fig.height=12}
EMP %>% 
  tax_glom(., Rank) %>% 
  transform_sample_counts(., function(x) x / sum(x) * 100) %>% 
  psmelt() %>%  
  mutate_if(is.character, as.factor) %>% 
  mutate(Taxa = fct_relevel(pull(., Rank), rev(unique(pull(Rank_order, Rank))))) %>%
  mutate(Taxa = fct_other(Taxa, drop = pull(Rank_order, Rank)[Rank_order$Abundance < 0.1], other_level = "Rare")) %>% 
  ggplot(aes(x = Taxa, y = Abundance)) +
  geom_violin(aes(group = interaction(Taxa, !!sym(Grouping_var))),
              scale = "width") +
  geom_point2(aes(colour = Taxa), 
              position = position_jitter(width = 0.2),
             alpha = 1/4,
             stroke = 0, 
             size = 2) +
  theme(axis.text = element_text(angle = 45.0), 
        axis.text.x = element_text(vjust = 1, hjust = 1) ) + 
  # scale_fill_locuszoom() +
  # scale_color_manual(values = pal("d3js")[c(3, 4, 2)]) +
  labs(x = NULL, y = "Abundance (%)", colour = Rank) + 
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1))) + 
  facet_wrap(~get(Grouping_var), ncol = 2) 
```

## Generate heat-tree plots using [Metacoder][3] [@foster_metacoder_2017]

```{r plot metacoder, fig.height = 30, fig.width = 18}
# Convert phyloseq to taxmap
EMP %>% 
  # transform_sample_counts(function(x) x / sum(x) ) %>% # not really needed
  parse_phyloseq() %>% 
  taxa::filter_taxa(., !taxon_ranks %in% c("Species", "Genus", "Family", "Order")) -> # tips will be Class
  EMP_taxmap

# Potentially drop out low reads
# EMP_taxmap$data$otu_table <- zero_low_counts(EMP_taxmap, "otu_table", min_count = 5)
# no_reads <- rowSums(EMP_taxmap$data$otu_table[, EMP_taxmap$data$sample_data$sample_id]) == 0
# EMP_taxmap <- filter_obs(EMP_taxmap, data = "tax_data", ! no_reads, drop_taxa = TRUE)

# Sum counts per group
EMP_taxmap$data$tax_counts <- calc_taxon_abund(EMP_taxmap, 
                                                      data = 'otu_table', 
                                                      cols = EMP_taxmap$data$sample_data$sample_id, 
                                                      groups = pull(EMP_taxmap$data$sample_data, Grouping_var))

# Generate a plot for each group
my_plots <- lapply(sort(unique(get_variable(EMP, Grouping_var))), function(group) {
  # do something with time point data
  set.seed(2020) # Make each plot layout look the same if using another layout
  EMP_taxmap %>% 
    heat_tree(.,
              node_label = taxon_names,
              node_size = n_obs,
              node_color = EMP_taxmap$data$tax_counts[[group]],
              node_size_axis_label = 'OTU count',
              node_color_axis_label = 'Sequence count',
              layout = "automatic",
              node_size_range = c(0.013, 0.04), # makes scale range the same in both
              title = group)
})

wrap_plots(my_plots, ncol = 3)
```

## Colophon
```{r colophon, eval=T}
sessioninfo::session_info() %>%
  details::details(
    summary = 'Current session info',
    open    = TRUE
  )
```

[1]: https://earthmicrobiome.org/
[2]: https://earthmicrobiome.org/protocols-and-standards/empo/
[3]: https://grunwaldlab.github.io/metacoder_documentation/